Panarchy: ripples of a boundary concept

### 2021-10-07 paper/paper_panarchy.Rmd
edits second round of review

To prepare the data for topic modeling, we constructed a document term matrix
with documents in rows (N=`r dim(dtm)[1]`) and words in columns
(N=`r dim(dtm)[2]`). Here our unit of analysis for the document are the
abstracts retrieved, and the matrix contains the count of the number of
occurrences of each word per abstract. We removed stop words (e.g. "the", "a")
and digits from the matrix, as well as a list of words that were over
represented in our data and are common in the scientific literature but are
unrelated to the papers' topics (e.g. "paper", "study", "aim"). The lexicon of
stop words has 1149 entries, curated by various research groups, and accessible
in the `R` language under the packages `tm` and `tidytext` [@R-topicmodels;
@R-tm; @R-NLP; @tm2008; @R-tidytext], for example by `tidytext::stop_words`. The
data was also lemmatized using the hash lemmas dataset available at
[\<http://www.lexiconista.com\>](http://www.lexiconista.com/). Lemmatization is
the process of simplifying words to their root lemmas (e.g. *teaching* and
*taught* are lemmatized to *teach*).

@@

We also identified what aspects of panarchy were most used in the papers. For
example whether there is emphasis on cross-scale interactions, or poverty and
rigidity traps. We used text annotations to capture potential criticisms as well
as the methods used. We did a preliminary scoping and coding of papers, and
through a iterative process we kept track of potential fields under reviewed by
our corpus, or simply absent of it. We coded papers until we reached conceptual
saturation, that is, additional papers were not proving additional categories of
the already coded. The code book is available at
[doi.org/10.6084/m9.figshare.13490919](http://doi.org/10.6084/m9.figshare.13490919).

@@

Several of the papers reviewed (81%) attempted to identify the adaptive cycle
either through qualitative or quantitative methods. Identification of adaptive
cycles has found applications in a wide range of disciplines and research
problems, from delimitation of periods in archaeology and anthropology
[@Redman:2003bd], to financial crises in Europe [@Castell:2020dg], to traffic
jams in China [@Zeng:2020hu]. Some recent empirical tests of the adaptive cycle
used innovative types of data and methods. For example, information transfer
methods based on entropy have been suggested to approximate relevant components
of a system and the empirical proxies of the axes within which the adaptive
cycle unfolds: potential, connectedness, and resilience [@Castell:2020dg].
Percolation methods combined with big data have been proven useful for testing
hypotheses of regime shifts in urban systems (originally proposed in Panarchy
[@Gunderson:2002vk]), and deriving the temporal and spatial scales at which the
adaptive cycle emerges [@Zeng:2020hu].

@@

The high level of conceptualization but lack of empirically grounded testing of
panarchy theory came across as one of the major limitations, given our sample of
papers reviewed (N=`r nrow(qual)`). What makes a theory a scientific theory is
that the plurality of scientific approaches (e.g. positivism, constructivism)
should have mechanisms in place to avoid confirmation bias, allowing the
emergence of alternative explanations, and identify the circumstances under
which a theory may fail [@Hansson:2017tr]. Many of the papers we read confirmed
but did not challenge the expectations that Panarchy put forward, though a few
of the most recent works did falsify some of the ideas proposed --e.g. the back
loop is faster that the front one [@Castell:2020dg]. We believe more work is
needed to better understand the circumstances where Panarchy works, and when it
falls short. In other words, to refine the scientific framework into a
scientific theory.

The criticism of over-conceptualization is not unique to the literature that
engages with Panarchy ideas. Here by over-conceptualization we refer to the
problem of scientific jargon, where different communities use slightly different
wordings to refer to the same phenomenon; reducing the chances that lessons from
a discipline percolates to another. An example from Panarchy is the endorsement
of ideas such as creative destruction (from economics) or the theory of
intermediate disturbances (from ecology). The latter has been debunked in the
ecological literature but might still been used within Panarchy thinking
[@Fox:2013br]. A recent review of sustainability science mapped the different
schools of thought that the discipline has developed over the last decades
[@Clark:2020jd] and reached a similar conclusion: too many conceptual frameworks
have been developed, but there have been too few empirical attempts to test the
frameworks against data and falsify hypotheses. The review also emphasizes the
problem of measurement and observation [@Clark:2020jd]. In the context of
panarchy, recent work has developed an information theory based approach that
enables the identification of adaptive cycles [@Castell:2020dg]. The authors
identify phases of the adaptive cycle in the European financial crises and in
grassland ecosystems, but fail to find support for the difference in speed
between the forward and backward loops originally proposed in Panarchy.

@@

Sustainability science is defined by the problems it addresses rather than the
disciplines it employs [@Anonymous:2016fv]. Our topic modeling results helped in
identifying some of the problems that the community is addressing and how they
have changed over time. Some of these problems include climate change (topic
24), urban development and disaster response (7), conservation and landscape
management trade-offs (5), natural resource management (9), food production (3),
policy tools and plans (16), community based management (13) or knowledge and
local social practices (25. We also observed topics closely related to methods
or approaches to study social-ecological systems such as modeling (19,21),
networks (6), or ecosystem services (15). Topic modeling can be used to further
investigate particular academic communities, identify key landmark papers in
certain topics, or identify the network of key collaborators and experts.

This paper aimed to study the use of panarchy related concepts and their
evolution since the publication of the book. Systematic literature reviews are
often restricted to a limited subset of the literature -- based on what is
readable on the time frame of a project -- of dozens to maybe hundreds of
papers. Qualitative analysis offers rich insights into the papers reviewed but
is limited by sample size in its ability to draw general conclusions. In our
analysis, the qualitative insights could be biased towards a non-random
selection of papers that aligned our research interests or highly cited review
papers. Topic modeling enabled us to complement the analysis and include all
papers indexed by major scientific databases. It has the advantage of
reproducibility and reduces sample bias, but offers limited insights into the
dimensions of panarchy used, methods, or criticisms. Here we show that both
methods combined are a powerful combination.

Our choice of methods does however have some limitations. We could not directly
compare the two methods since they do not share a common categorical axis on
which they can be measured back to back. The topics gave us a general but
shallow overview, while the document analysis provide deep insights on
particular work but limited generalizations. We also, for example, ignored rich
areas of social science research (e.g. Anthropology) that is published in books
and book chapters, that are not archived by scientific search engines.
Additionally, a more inductive approach to the qualitative coding could have
given us different insights on the literature outside the pre-defined categories
we initially shared with the Resilience Alliance.

>>>>

To prepare the data for topic modeling, we constructed a document term matrix
with documents in rows (N=`r dim(dtm)[1]`) and words in columns
(N=`r dim(dtm)[2]`). Here our unit of analysis for the document are the
abstracts retrieved, and the matrix contains the count of the number of
occurrences of each word per abstract. We removed stop words (e.g. "the", "a")
and digits from the matrix, as well as a list of words that were over
represented in our data and are common in the scientific literature but are
unrelated to the papers' topics (e.g. "paper", "study", "aim"). The lexicon of
stop words has 1149 entries, curated by various research groups, and accessible
in the `R` language under the packages `tm` and `tidytext` [@R-topicmodels;
@R-tm; @R-NLP; @tm2008; @R-tidytext], for example by `tidytext::stop_words`. The
data was also lemmatized using the hash lemmas dataset available at
[\<http://www.lexiconista.com>](http://www.lexiconista.com/). Lemmatization is
the process of simplifying words to their root lemmas (e.g. *teaching* and
*taught* are lemmatized to *teach*).

@@

We also identified what aspects of panarchy were most used in the papers. For
example whether there is emphasis on cross-scale interactions, or poverty and
rigidity traps. We used text annotations to capture potential criticisms as well
as the methods used. We did a preliminary scoping and coding of papers, and
through an iterative process we kept track of potential fields under review by
our corpus, or simply absent of it. We coded papers until we reached conceptual
saturation, that is, additional papers were not proving additional categories of
the already coded. The code book is available at
[doi.org/10.6084/m9.figshare.13490919](http://doi.org/10.6084/m9.figshare.13490919).

@@

Several of the papers reviewed (81%) attempted to identify the adaptive cycle
either through qualitative or quantitative methods. Identification of adaptive
cycles has found applications in a wide range of disciplines and research
problems, from delimitation of periods in archaeology and anthropology
[@Redman:2003bd], systemic interactions between forest-related sectors in Sweden [@moen2010interlocking], financial crises in Europe [@Castell:2020dg], to
traffic jams in China [@Zeng:2020hu]. The adaptive cycle has been used as a model to structure rich empirical information about participation in adaptive management and, based on that analysis, to develop recommendations for more flexible stakeholder engagement [@stringer2006unpacking]. The adaptive cycle has also been used to provide a more mechanistic description of the history of a variety of specific social-ecological systems, including the Galapagos Islands [@gonzalez2008rethinking], the Western Australian Agricultural Region [@allison2004resilience], the Erhai Lake watershed in southwest China [@dearing2008landscape], and the cities of Cambridge and Swansea in the UK [@simmie2010economic]. The adaptive cycle has also been used to describe more general social and ecological processes, such as lake eutrophication [@carpenter2001metaphor] and the development of regional industrial clusters [@martin2011conceptualizing].

Recent empirical tests of the
adaptive cycle used innovative types of data and methods. For example,
information transfer methods based on entropy have been suggested to approximate
relevant components of a system and the empirical proxies of the axes within
which the adaptive cycle unfolds: potential, connectedness, and resilience
[@Castell:2020dg]. Percolation methods combined with big data have been proven
useful for testing hypotheses of regime shifts in urban systems (originally
proposed in Panarchy [@Gunderson:2002vk]), and deriving the temporal and
spatial scales at which the adaptive cycle emerges [@Zeng:2020hu].

@@

The high level of conceptualization but lack of empirically grounded testing of
panarchy theory came across as one of the major limitations, given our sample of
papers reviewed (N=`r nrow(qual)`). What makes a theory a scientific theory is
that the plurality of scientific approaches (e.g. positivism, constructivism)
should have mechanisms in place to avoid confirmation bias, allowing the
emergence of alternative explanations, and identify the circumstances under
which a theory may fail [@Hansson:2017tr]. Many of the papers we read
confirmed but did not challenge the expectations that Panarchy put forward,
though a few of the most recent works did falsify some of the ideas proposed
--e.g. the back loop is faster than the front one [@Castell:2020dg]. We
believe more work is needed to better understand the circumstances where
Panarchy works, and when it falls short. In other words, to refine the
scientific framework into a scientific theory.

The criticism of over-conceptualization is not unique to the literature that
engages with Panarchy ideas. Here by over-conceptualization we refer to the
problem of scientific jargon, where different communities use slightly different
wordings to refer to the same phenomenon; reducing the chances that lessons from
a discipline percolates to another. An example from Panarchy is the endorsement
of ideas such as creative destruction (from economics) or the theory of
intermediate disturbances (from ecology). The latter has been debunked in the
ecological literature but might still have been used within Panarchy thinking
[@Fox:2013br]. A recent review of sustainability science mapped the different
schools of thought that the discipline has developed over the last decades
[@Clark:2020jd] and reached a similar conclusion: too many conceptual
frameworks have been developed, but there have been too few empirical attempts
to test the frameworks against data and falsify hypotheses. The review also
emphasizes the problem of measurement and observation [@Clark:2020jd]. In the
context of panarchy, recent work has developed an information theory based
approach that enables the identification of adaptive cycles
[@Castell:2020dg]. The authors identify phases of the adaptive cycle in the
European financial crises and in grassland ecosystems, but fail to find support
for the difference in speed between the forward and backward loops originally
proposed in Panarchy.

@@

Sustainability science is defined by the problems it addresses rather than the
disciplines it employs [@Anonymous:2016fv]. Our topic modeling results helped
in identifying some of the problems that the community is addressing and how
they have changed over time. Some of these problems include climate change
(topic 24), urban development and disaster response (7), conservation and
landscape management trade-offs (5), natural resource management (9), food
production (3), policy tools and plans (16), community based management (13) or
knowledge and local social practices (25). We also observed topics closely
related to methods or approaches to study social-ecological systems such as
modeling (19,21), networks (6), or ecosystem services (15). Topic modeling can
be used to further investigate particular academic communities, identify key
landmark papers in certain topics, or identify the network of key collaborators
and experts.

This paper aimed to study the use of panarchy related concepts and their
evolution since the publication of the book. Systematic literature reviews are
often restricted to a limited subset of the literature -- based on what is
readable on the time frame of a project -- of dozens to maybe hundreds of
papers. Qualitative analysis offers rich insights into the papers reviewed but
is limited by sample size in its ability to draw general conclusions. In our
analysis, the qualitative insights could be biased towards a non-random
selection of papers that aligned with our research interests or highly cited review
papers. Topic modeling enabled us to complement the analysis and include all
papers indexed by major scientific databases. It has the advantage of
reproducibility and reduces sample bias, but offers limited insights into the
dimensions of panarchy used, methods, or criticisms. Here we show that both
methods combined are a powerful combination.

Our choice of methods does however have some limitations. We could not directly
compare the two methods since they do not share a common categorical axis on
which they can be measured back to back. The topics gave us a general but
shallow overview, while the document analysis provide deep insights on
particular work but limited generalizations. We also, for example, ignored rich
areas of social science research (e.g. Anthropology) often published in books
and book chapters, that are not archived by scientific search engines.
Additionally, a more inductive approach to the qualitative coding could have
given us different insights on the literature outside the pre-defined categories
we initially shared with the Resilience Alliance.
